\Title

### 2011-03-24 masters_thesis/proposal/proposal.tex
Finished my proposal and got it approved by Ventura

My honors thesis proposed and developed this basic approach to parallelization
in PSO.  This masters thesis will complete the picture of speculative methods
in particle swarm optimization.  First, we present a mathematical proof of
correctness of our methods, given in Section~\ref{sec:proof} as demonstration
that what we propose is in fact mathematically equivalent to standard
parallelization methods in PSO.  Second, we propose pruning the set of possible
speculative evaluations to better use available processors.  We motivate the
pruning by analyzing the statistics of which branches are commonly taken in
PSO, using those statistics to prune all but the most likely branches.
Further, once the idea of pruning is established, we view the set of all
possible speculative evaulations as an infinite tree, not confining ourselves
to speculating only about the next iteration.  We then experiment with
speculating down likely paths three or four iterations ahead, and show that
this by far outperforms simple one-iteration speculative evaluation.  And
lastly, as my honors thesis used only a few experiments on three benchmark
functions, we give a better picture of the performance of speculative
approaches with much more extensive experimentation.

\subsection{Refactoring the PSO Equations}
\label{sec:proof}

To perform two iterations at a time we must first refactor PSO such that the
determination of the value of the objective function is separate from the rest
of the computation.  For simplicity, this discussion will describe the case
where PSO is performing function minimization using the Ring topology.  In this
example, each particle has two neighbors, the ``right neighbor'' and ``left
neighbor,'' whose positions are represented as $\rightn$ and $\leftn$
respectively.  Though we will only describe the case of the Ring topology, this
method can easily be extended to arbitrary topologies (and in fact our
implementation of this algorithm works with arbitrary topologies; we simply
only discuss the Ring case here as that greatly simplifies our notation).

The refactoring hinges on the idea that once the random coefficients
$\pURand_{t}$ and $\nURand_{t}$ are drawn, there are only a few possible new
positions, or updates, for $\nbest$ and $\pbest$.  For the Ring topology there
are 7 possible update cases, identified in Table~\ref{tab:evals}.  We label
each case with an identifier referring to the source of the update: a minus
sign ($-$) represents no update, $L$ represents an update to $\nbest$ coming
from the left neighbor, $R$ represents an update to $\nbest$ coming from the
right neighbor, and $S$ represents an update to either $\pbest$ or $\nbest$
coming from the particle itself.  As an example, $\casexn$ refers to the case
that the particle finds a new personal best, but neither it nor its neighbors
find a position that updated its neighborhood best.  In the equations that
follow, we refer to an unspecified update case as $\casegen$, and to the set of
cases collectively as $\caseset$.

\begin{table}
  \caption{All possible updates for a particle with two neighbors}
  \label{tab:evals}
  \centering
  \begin{tabular}{lcc}
	Identifier&Source of $\pbest$ update&Source of $\nbest$ update\\
	\hline
	\hline
	$\casepn$&No update&No update\\
	\hline
	$\casepl$&No update&Left Neighbor\\
	\hline
	$\casepr$&No update&Right Neighbor\\
	\hline
	$\casexn$&Self&No update\\
	\hline
	$\casexl$&Self&Left Neighbor\\
	\hline
	$\casexr$&Self&Right Neighbor\\
	\hline
	$\casexx$&Self&Self\\
	\hline
  \end{tabular}
\end{table}

In order to incorporate the determination of which case occurs into the
position and velocity update equations, we introduce an indicator function
$\indic_{t+1}^{\casegen}$ for each case $\casegen \in \caseset$.  When
$\casegen$ corresponds to the case actually taken by PSO,
$\indic_{t+1}^{\casegen}$ evaluates to 1; otherwise it evaluates to 0.  We can
then sum over all of the cases, and the indicator function will make all of the
terms drop to zero except for the case that actually occurs.  For example, the
indicator function for the specific case $\casexn$ (which, as is shown in
Table~\ref{tab:evals}, means that the particle's personal best was updated, but
its neighborhood best was not) can be written as follows:

\begin{align}
  \nonumber
	\indic_{t+1}^{\casexn} & (\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1}))= \\
  \label{eq:deficasexn}
	&\begin{cases}
	   1 & \text{if} \ \ofunc(\ppos_{t}) < \ofunc(\pbest_{t-1}) \\
	   &\quad \text{and} \quad \ofunc(\nbest_{t-1}) < \ofunc(\ppos_{t}) \\
	   &\quad \text{and} \quad \ofunc(\nbest_{t-1}) < \ofunc(\leftn_{t}) \\
	   &\quad \text{and} \quad \ofunc(\nbest_{t-1}) < \ofunc(\rightn_{t}) \\
	   0 & \text{otherwise}
	\end{cases}
\end{align}

For each case $\casegen \in \caseset$, there is also a corresponding
velocity update function $\specvel_{t+1}^{\casegen}$.  When the case is
known, the specific values of $\pbest_t$ and $\nbest_t$ may be substituted
directly into~\eqref{eq:velupdate}.  For example, in case $\casexn$,
$\pbest_{t}=\ppos_{t}$, as \pbest was updated by the particle's current
position, and $\nbest_{t}=\nbest_{t-1}$, as $\nbest$ was not updated at
iteration $t$:
\begin{align}
\nonumber
	\specvel_{t+1}^{\casexn} & (\pvel_t, \ppos_{t}, \leftn_{t}, \rightn_{t},
	\pbest_{t-1}, \nbest_{t-1}, \pURand_{t}, \nURand_{t}) \\
\label{eq:defvcasexn}
		&= \constriction \bigl[ \pvel_{t} +
			\pcoeff\pURand_{t}\otimes(\ppos_{t} - \ppos_{t})
			+ \ncoeff\nURand_{t}\otimes(\nbest_{t-1} -
			\ppos_{t}) \bigr]
\end{align}

In the same way we can create notation for the position update function by
substituting into~\eqref{eq:posupdate}.  For compactness, we will drop the
parameters to $\specvel_{t+1}^{\casegen}$ since they can be inferred from the
subscripts.
\begin{align}
\label{eq:defpcasegen}
	\specpos_{t+1}^{\casegen} & (\ppos_{t}, \pvel_{t}, \leftn_{t},
	\rightn_{t} ,\pbest_{t-1} ,\nbest_{t-1}, \pURand_{t}, \nURand_{t})
	= \ppos_{t} + \specvel_{t+1}^{\casegen}
\end{align}

With this notation we can re-write the original PSO velocity
equation~\eqref{eq:velupdate}, introducing our sum over cases with the
indicator functions.  Again, we represent the indicator functions and velocity
functions without the parameters for compactness.  The equation becomes:
\begin{align}
\nonumber
	\pvel_{t+1} &=
		\constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t})
			+ \ncoeff\nURand_{t}\otimes(\nbest_{t} -
			\ppos_{t}) \bigr] \\
\nonumber
	&= \sum_{c \in \caseset} \indic_{t+1}^{c} \ \constriction \bigl[ \pvel_t
			+ \pcoeff\pURand_{t}\otimes(\pbest_{t} - \ppos_{t})
			+ \ncoeff\nURand_{t}\otimes(\nbest_{t} -
			\ppos_{t}) \bigr]  \\
\label{eq:vel2update}
	&= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specvel_{t+1}^{c} 
\end{align}

Similarly, the position update equation~\eqref{eq:posupdate} becomes:
\begin{align}
\label{eq:pos2update}
	\ppos_{t+1} &= \ppos_{t} + \pvel_{t+1}
	= \sum_{c \in \caseset} \ \indic_{t+1}^{c} \ \specpos_{t+1}^{c} 
\end{align}

The value of the objective function at $\ppos_{t+1}$ is given by:
\begin{align}
\label{eq:val2update}
	\ofunc (\ppos_{t+1}) = \sum_{c \in \caseset} \ \indic_{t+1}^{c}
	\ \ofunc(\specpos_{t+1}^{c})
\end{align}

Returning our attention to the computation of $\ppos_{t+1}$ in
\eqref{eq:pos2update} and writing it with the parameters which were omitted
above, we obtain:
\begin{align}
\nonumber
  \ppos_{t+1} = \sum_{c \in \caseset}
	&\indic_{t+1}^{c}(\ofunc ( \ppos_{t} ) ,\ofunc(\leftn_{t}),
	\ofunc(\rightn_{t}) ,\ofunc(\pbest_{t-1}) ,\ofunc(\nbest_{t-1})) \\
\label{eq:val2updatelong}
	& \specpos_{t+1}^{c}(\ppos_{t},\pvel_{t},\leftn_{t},\rightn_{t},
	\pbest_{t-1},\nbest_{t-1},\pURand_{t}, \nURand_{t})
\end{align}

In this form the important point to notice is that there are only $7$ values
(for this Ring topology) in the set $\{\specpos_{t+1}^{\casegen}: \casegen \in
\caseset\}$ and that none of them depend upon $f(\ppos_t)$ or any other
objective function evaluation at iteration $t$.  Note also that while there are
random numbers in the equation, they are assumed fixed once drawn for any
particular particle at a specific iteration.  Thus PSO has been refactored such
that the algorithm can begin computing all $7$ of the objective function
evaluations potentially needed in iteration $t+1$ \emph{before} $f(\ppos_t)$ is
computed.  Once the evaluation of $f(\ppos_{t})$ is completed for all particles
only one of the indicator functions $\indic_{t+1}^{\casegen}$ will be set to 1;
hence only one of the positions $\specpos_{t+1}^\casegen$ will be kept.

Although this speculative approach computes $\ofunc(\specpos_{t+1}^{\casegen})$
for all $\casegen \in \caseset$, even those for which $\indic_{t+1}^{\casegen}
= 0$, these extra computations will be ignored, and might just as well never
have been computed.  We call the set of computations
$\{\ofunc(\specpos_{t+1}^{c}) : \casegen \in \caseset\}$ ``speculative
children'' because only one of them will be needed.

@@

For each function, we will use the topology that is generally considered to be
best for optimizing that particular function.  When the Complete topology is
considered best, we will instead use a Random topology that we developed that
mimicks the behavior of Complete while being more suited for parallel
computation~\cite{mcnabb-2009-large-particle-swarms}.

The functions we will use in our comparisons are these common benchmark
functions: Bohachevsky, Griewank, Rastrigin, Sphere, and Schwefel 2.21.  We
will run each function with 20, 50, and 500 dimensions to test the scalability
of our methods up to higher dimensions.  We will also test our methods on a
more typical ``real-world'' problem, that of training a computationally
intensive machine learning algorithm with a large amount of data.  We will
compare our speculative methods to previous parallelization techniques,
including a naive parallelization of just increasing the swarm size, and a
subswarms implementation that uses several smaller, independent swarms.

@@

\bibliography{../../../bib/pso/general/bib,%
../../../bib/pso/parallel/bib,%
../../../bib/pso/topology/bib,%
../../../bib/pso/fips/bib,%
../../../bib/aml/bib,%
../../../bib/parallel/bib,%
../../../bib/nfl/bib%
}

>>>>

My honors thesis proposed and developed this basic approach to parallelization
in PSO.  This masters thesis will complete the picture of speculative methods
in particle swarm optimization.  First, we present a mathematical proof of
correctness of our methods as demonstration that what we propose is in fact
mathematically equivalent to standard parallelization methods in PSO.  Second,
we propose pruning the set of possible speculative evaluations to better use
available processors.  We motivate the pruning by analyzing the statistics of
which branches are commonly taken in PSO, using those statistics to prune all
but the most likely branches.  Further, once the idea of pruning is
established, we view the set of all possible speculative evaluations as an
infinite tree, not confining ourselves to speculating only about the next
iteration.  We then experiment with speculating down likely paths three or four
iterations ahead, and show that this by far outperforms simple one-iteration
speculative evaluation.  And lastly, as my honors thesis used only a few
experiments on three benchmark functions, we give a better picture of the
performance of speculative approaches with much more extensive experimentation.

\subsection{Proof}

Some readers of my honors thesis had a hard time understanding our methods,
thinking that we were changing the PSO algorithm when performing speculative
evaluation.  To alleviate any such concerns and provide a strong theoretical
background for speculative approaches to parallelization in PSO, we will
provide a mathematical proof that PSO can be decomposed in such a way as to
exactly reproduce the behavior of the original algorithm.

\subsection{Statistically Motivated Pruning}

The key idea in this proposal is that of decreasing the number of speculative
evaluations needing to be done by pruning away insignificant branches from the
set of all possible evaluations.  In my honors thesis I found that exactly
reproducing PSO two iterations at a time required too many extra evaluations to
give competitive results in most cases.  To fix this problem, I will find ways
to still speculate about future positions without actually evaluating all of
the possible options.  This should improve the performance of the algorithm.

Pruning the set of speculative evaluations necessitates a mechanism for
deciding which branches are desirable.  A na\"ive approach to pruning would be
to try the last branch taken by each particle.  Preliminary experimentation
with this method has shown it to be ineffective.  A smarter approach would be
to examine the statistical behavior of the branches that PSO takes and use that
analysis as motivation for deciding which branches to prune.  We will collect
statistics on the branching behavior of PSO across several functions and
topologies.  We will then use those statistics to perform intelligent pruning
of the set of speculative evaluations.

\subsection{Speculating Several Iterations Ahead}

Speculating one iteration ahead proved very costly in terms of the number of
extra evaluations required.  Speculating two iterations ahead would require the
square of the number of extra evaluations of speculating a single iteration
ahead.  While exactly speculating two iterations ahead would be far too costly
to be practical, when combined with statistical pruning we can experiment with
speculating along statistically promising lines several iterations ahead.
Preliminary experimentation shows that this can be incredibly effective,
improving the time to completion by a factor of 2 or more, though it also seems
prone to premature convergence.  This thesis will further explore the idea of
speculating several iterations ahead.

@@

For each function, we will use the topology that is generally considered to be
best for optimizing that particular function, as described by Bratton and
Kennedy~\cite{bratton-2007-defining-a-standard-for-pso}.  When the Complete
topology is considered best, we will instead use a Random topology that we
developed that mimics the behavior of Complete while being more suited for
parallel computation~\cite{mcnabb-2009-large-particle-swarms}.

The functions we will use in our comparisons are these common benchmark
functions: Bohachevsky, Griewank, Rastrigin, Sphere, and Schwefel 2.21.  We
will run each function with 20, 50, and 500 dimensions to test the scalability
of our methods up to higher dimensions.  We will also test our methods on a
more typical ``real-world'' problem, that of fitting a model to a large amount
of data.  While we will run tests on a complex crystallography problem in
collaboration in Dr. Branton Campbell in the physics department, we will
present our results on a similar but more well known problem.  In particular,
we will train a radial basis function on a set of 10,000 data points.  This is
a similar problem to Dr. Campbell's, in that both seek to match model
parameters to observed data, but it is easier to describe and quicker to
experiment with.  Dr. Campbell also has yet to publish his results and we do
not wish to publish about his problem before he does.

We will compare our speculative methods to previous parallelization techniques,
including a naive parallelization of just increasing the swarm size, and a
subswarms implementation that uses several smaller, independent swarms.

@@

\bibliography{bib}
